\begin{acknowledgements}

It feels a little strange addressing the task of writing this section. Almost the entire thesis is expected to be structured, analytical and technical. Taking some time to examine how you're feeling about the whole process and who you have to be thankful for can be difficult for physicists, I'd like to present a quantitative analysis of those who've helped me through but I think that might be a force of habit. Instead, I'll try to talk from the heart and have a little think about those who've got me to this stage.

First, to my supervisors Lesley and Scott. Your generosity, technical advice and personal guidance over the past 4 years has been unrivalled. Lesley - your belief in me, kindness and understanding (particularly when I was struggling) has had more positive affect on my life than you probably know. Scott - your laughter, energy and enthusiasm for research is truly infectious. I can safely say I wouldn't have made it anywhere near this point had it not been for the two of you. I couldn't have asked for two better mentors - thank you both.

To office 113 (and neighbours) - Matt, Helena, Milan, Kai, Beatrice  

To Josh and Jorge - thank you for unnecessarily long coffee breaks and putting the world to right during them.

To my Mum and Dad. For unwavering support and belief (and bringing Cassie to visit).

To Ella, Louis and Hamish. For being amazing housemates, suffering through the low points of a PhD together, making outstanding pasta and enjoying a good cup of herbal tea. 

To Claire. You probably need your own chapter for this. For travelling from Edinburgh to see me, for amazing trips to the US, for picking me up when I was low and reminding me that I can always be a weatherman. 

To Cameron, Josh and Ben. For making the years studying physics at Edinburgh some of the best I've had. For allowing me to continue in physics and always giving me the confidence to keep going. You're all wonderful people who probably don't realise how much you've helped.

To Rip R and Chet M. 


\end{acknowledgements}




%%% Local Variables:
%%% mode: latex
%%% TeX-master: "thesis"
%%% End:
