\chapter{The role of vertical structure in QBO teleconnections}
\label{cha:deepQBO}

\section{Introduction}
\label{sec:deepQBO-introduction}
The QBO is typically defined by the equatorial ZMZW at a single level in the mid-stratosphere. The 50 hPa level is usually used for NH observational studies \citep{Baldwin2001, Baldwin98} but some studies have also noted the importance of characterising the vertical structure of the QBO \citep{Fraedrih1993, Wallace1993,  Baldwin98,  Dunkerton2017, Gray2018, andrewsObserved2019}. In an observational-based study \cite{Gray2018} find an enhanced association between the QBO and polar vortex when a metric incorporating the vertical coherence of equatorial winds via empirical orthogonal functions is utilised \citep{verena2016a}. In a model-based study \cite{Andrews2019} introduce a similar but simpler methodology by defining the QBO as the average ZMZW between two vertical levels, which preferentially selects time intervals that display a vertically coherent QBO phase between the specified levels. The same study reports enhanced responses to the QBO from the NAO and Arctic Oscillation (AO) when such vertically integrated metrics are used to define QBO phase compared to a single level definition. While these studies highlight statistical associations between vertical QBO metrics and the vortex, the importance of vertical coherence in HT strength has not been tested explicitly. Furthermore, influence mechanisms are not well understood. 


\section{Model Configuration}

To explicitly test the effect of QBO vertical coherence on teleconnections we utilise a set of custom designed simulations which utilises a modified version of the UKESM GCM analysed in chapters 3 and 4. As with the pi-control simulation examined so far, the model component for the atmosphere is GA7.1, GHG concentrations are set to 1850 levels (see section \ref{sec:model_config}) and volcanic eruptions are absent (however a constant background volcanic aerosol level is imposed). The primary difference between the pi-control and the configuration utilised here is the experiments run with no coupled Ocean component. Instead, SSTs are prescribed using a seasonally invariant climatological field obtained from an interval of the the pi-control (year numbers 96-126). Sea Ice extents are also prescribed to climatological values using the same interval of the pi-control and full documentation for this type of timeslice experiment is provided in 

We choose to impose SST and sea ice conditions in our experiments as this removes contributions to atmospheric circulation from interannual to decadal surface modes of variability such as ENSO and the AL. This in turn isolates the effect of the QBO further as any changes in QBO-vortex and QBO-surface interactions in the simulations are not due to coupling with SSTs and other surface boundary conditions. All simulations are initialised using conditions from the start of the 30 year interval of the pi-control used to calculate SST and SI climatologies (year number 96). Analysis chapters 4 and 5 showed that this year does not lie in an interval in which the QBO or the vortex exhibits significant multi-decadal oscillations. As a result, the initial conditions for the simulation are less likely samples the QBO and vortex in a randomly chosen state.

\section*{Nudging}
In order to impose the vertical structure of the QBO in our sensitivity experiments we employ a method to constrain atmospheric conditions in a GCM simulation known as nudging. In this method, an atmospheric quantity, $Y$, is constrained via Newtonian-relaxation which pushes the model state for $Y$ towards some imposed value at each model timestep such that

\begin{equation} \label{eq:nudging}
\Delta Y = G \Delta t (Y_{imposed} - Y_{model}), 
\end{equation}

where $Y_{imposed}$ is the state of $Y$ from the imposed field, $Y_{model}$ is the value of $Y$ the model has evolved to since the last timestep (when it was last nudged), $\Delta Y$ is the change in $Y$ applied by the nudging scheme and $G$ is the relaxation parameter. $G$ is a measure of the strength of the relaxation towards $Y_{imposed}$ at each timestep and can also be expressed as $G = \frac{1}{\tau}$ where $\tau$ is known as the relaxation timescale. When choosing the relaxation timescale, a balance is required between constraining the model sufficiently to the imposed state (I.E. a short enough $\tau$ such that the model $Y$ avoids evolving on its own) and avoiding the introduction of large gradients and (a large enough $\tau$ to avoid model instability). The MetOffice Unified model, of which the model configurations presented here are a part, typically uses a relaxation timescale of 6 hours at all vertical levels. This has been shown to produce good quality nudging results in a range of studies employing such a method despite the different radiative timescales present across atmospheric levels (Brown et al. 2020). In our experiments we to nudge the zonal wind, $U$, towards an idealised version of the QBO wind fields expressed as 

\begin{equation} \label{eq:imposed_U}
U_{qbo}(t, z) = F(z) sin(\frac{2\pi (t + zs)}{period}),
\end{equation}

Where $s$ is a parameter which sets the rate of phase shift with descending height. We set $s$ to xxxx and the For the deep QBO experiment (that which displays a high degree of vertical coherence) and xxx for the shallow QBO to ensure a large phase shift with height (and therefore less vertical coherence). $F(z)$ is a momentum forcing vertical profile (seen in figure xxx) applied to reproduce the variation in magnitude of the QBO with height. Following the methodology of Pascoe et al., we use an $F(z)$ which takes the form of a Weibull function given by

\begin{equation} \label{eq:vertical_profile}
F(z) = R_D \frac{1}{\gamma \beta^\alpha}  Z_D^{\alpha-1}  e^{-Z_D/\beta},
\end{equation}

where the parameters are set to $\alpha = 3.5, \beta = 0.27, \gamma = 3.2335, R_D = 40.0 ms^{-1}, Z_{D} = (z - 15)/20$. The timeseries generated for both experiments are shown in figure xxxxa, they highlight the differences between idealised QBOs: The deep QBO winds exhibits the same sign of wind throughout the middle atmosphere for the majority of the series while the shallow QBO show different phases between the upper and middle stratosphere. 